---
title: "Quosure Data Structure"
author: "Dan Sullivan"
date: "12/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

>Is a tree with two trunks a single tree?


##Introduction
This tutorial is about the structure of a quosure. It's purpose is to provide a 
foundation for another tutorial that will follow it about using quosures to 
*program on the langauge* like what is done by many of the functions in the
the tidyverse.

Some of the functions in this tutorial have the prefix "qtls_". They are not
available yet but will be as part of the tutorial that follows this one.

## Basics
A quosure object is a wrapper around an expression. The `typeof` a quosure is
`language` meaning a quosure is a language object. 

```{r typeof}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(quotools))
# make a quosure.. we'll be using this
# for some of what folows
q <- rlang::quo(a * b + c * d)
typeof(q)
```

In R the fact that a quosure is a `language` object means that it has certain
behaviors. For example the `rlang` package has a number of
functions that have a prefix of `lang_`. These functions can only be used on
language objects.

The `typeof` function, however, doesn't tell you everything about an 
object. In R a single object may mimic the behavior of many other
types objects.


```{r interfaces}
# show all the types of objects q mimics
qtls_what_is_it(q)

```
This shows that a quosure mimics a `formula` object, so you can treat it like a
formula. For example, any of the functions in the `rlang` package
that are prefixed with `f_` can be used with a `formula` object. 

A formula has a right and a left hand side. The right hand side, or rhs, is used
to store the expression that the quosure has wrapped.

```{r rhs}
qrhs <- rlang::f_rhs(q)
qrhs
```

Here you see the expression the quosure is wrapping in the rhs of the quosure.

The typeof the right hand side (rhs) of a quosure is  `language` and it also 
mimics many other types of objects.

```{r rhs2}
typeof(qrhs)
qtls_what_is_it(qrhs)
```
Here we see that the rhs mimics an `expr` object so you can treat it as an
expression. This means you can use any of the functions in the `rlang`
package with the prefix `expr_` with it.

#car and cdr

A language object, like the expression we are working on here,
has some important properties called car and cdr (pronounced sort of like "cudder").
Don't worry about the strange
names, they come from R's lisp heritage but their original meanings don't
apply here.

These properties can contain, at least in theory, any kind of objects, even
`list` objects. The `rlang` package gives you access to them.

```{r lists}
# get the car object from the rhs of quosure we are working on
qcar <- rlang::node_car(qrhs)
# find out what it is
typeof(qcar)
# get the cdr
qcdr <- rlang::node_cdr(qrhs)
# find out what it is
typeof(qcdr)
```

Here, in this example, we see that the car is a `symbol` and the cdr is a
`pairlist`. You probably haven't seen a `pairlist` before because they are
not used, at least not now, in R
applications. But for all practical purposes they are the same thing as a `list`
object.

The list in the cdr can contain, again in theory, any kinds of objects, 
including expression objects.

Given a cdr list with expression objects in it then via recursion this can
be used to build a tree. In fact this would make a quosure look like a data
structure we call a tree where each object in the tree contains a list
of its children.

We see these kinds of tree data structures all over the place. Examples
are the W3C DOM (object model of a web page and XML), the XPath model for XML,
scopes in a programming language, JSON and on and on.

#No Ordinary Tree

But we kind of left the car behind. The car can also contain any kind of object
including a list. But an object in a tree data structure only has a single list of
children. Maybe, by convention, R doesn't ever put a list in the car? If that
were the case than a quosure would be a classic tree data structure.

However that convention is not followed by R as we can see here...

```{r mulitplecar}
# anonymous function defined and executed in place
qf <- rlang::quo(function(x){x}(6) )
# get is rhs, i.e. the expression
qfrhs <- rlang::f_rhs(qf)
# get the cdr for that expresion
qfcdr <- rlang::node_cdr(qfrhs)
length(qfcdr)
# this tells us the cdr acts as list with 3 objects in it
# get the second object from this list
n2 <- qfcdr[[2]]
typeof(n2)
# n2 is a language object
# so you can get a car from it
n2car <- rlang::node_car(n2)
length(n2)
# this tells us that n2 acts as list with 2 objects in it
```

So sometimes a car will have a list of children in it. That means our quosure
data structure is not a tree, certainly not in the conventional sense.

But when we say a data structure
is a tree it is just so that we can have handy visualization to refer to in order
to grasp the its overall structure without having to think about all the details of it.
But in the end objects have properties and we make use of them in any way we
wish to fulfill the requirements of an application. 

Since the purpose of a tree is to give us a visualization that will help
us understand a data structure without having to look at every detail of it, we
can use a convention that says the single list of a language object is the
concatenation of the car and cdr lists, in that order.

And as long as each
object in the visualization shows whether it came from a the car or the cdr
you will have a good overall picture of the data structure.
With that in mind here is a visualization we can get 
for that anonymous, executed in place function we just looked at.

```{r mulitreee}
suppressPackageStartupMessages(library(DiagrammeR))
qftab <- qtls_make_rlang_table(qf)
g <- qtls_plot_model(qftab)
DiagrammeR::render_graph(g, layout="tree")
```

The visualization is a little messy but each node in the tree is
represented by a circle. The first line in each node is some descriptive
information about the object node represents. The second line tells us if 
that node came from the car or cdr of it's parent. So it's pretty
straightforward to see the overall  structure of a quosure.

If you look at the three nodes at the bottom of the tree can see that
two of those nodes came from the car of their parent.

There are other ways to visualize a tree that can present a lot more information
without being so cluttered and make it a lot easier to understand what is
going on inside a quosure. We'll see one of those visualizations in the
next tutorial.

#Summary
A quosure wraps an expression by turning into a one sided formula object, 
which the quosure can mimic.

The rhs of the quosure is a language object and stores the expression that the
quosure has wrapped.

The `typeof` the rhs is `langauge`. That means that it has`car` and `cdr`
properties. The `rlang` package has a number of functions the can be
used to work with those properties. You can use those functions to *ferret out*
a tree structures for a quosure.

A car or a cdr might be a solitary object or it might be a list of objects. 

We can represent a quosure as a tree if 
we use a convention that says that the car and cdr are lists they are concatenated
to make a single list.

There is important kind of tree called an abstract syntax tree or AST. The
tree structure in a quosure is **not** an AST. An AST is important because
it is a standard way many language implementations use to make, in effect, 
a list of directions that can be used to evaluate an expression. Fortunately
an AST can be extracted from the "double list" tree inside a quosure.

We'll see more on AST's and other topics in the tutorial on using quosures
to *program on the language* that follows this one.






